---
title: "Data"
author: "Your Name"
date: "2024-06-01"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(arrow)
library(lubridate)
library(skimr)
source("code/load_data.R")  # for sampling code if needed
```

## Load the data

The data for this case study is stored in a Parquet file. We will use the `arrow` package to read the data into R.
```{r load-data}
# List the parquet files
parquet_files <- list.files(
  "Unit 14 and 15 Case Study 2", 
  pattern = "yellow_tripdata_2025.*\\.parquet$",
  full.names = TRUE
)

# Open all files as a single Arrow Dataset (Lazy loading)
ds <- open_dataset(parquet_files, format = "parquet")

# Check what we have
ds$schema
nrow(ds)  # ~48.7 million rows
```

## Sampling the data

We'll sample from this dataset to create a manageable subset for our analysis using stratified random sampling to ensure we have all types of trips represented.

```{r}
set.seed(6306)  # for reproducibility
```

One confounding variable in tip anaylsis is cash or credit. Cash is largely determined by how much a passanger has in their pocket and we cannot assume it is reliably recorded due taxes prior to July 4, 2025 and manual entry methods. we will assume cash tipping is too confounding to model. 

We will pull two samples to justify our assumption and continue our analysis. One including cash tips and one with only credit card tips.

```{r}
# 2. Sampling + feature engineering (filter to credit-card trips first)
#    Use map_batches for memory-safe random sampling on huge data
#    Adjust the proportion to hit your target sample size (e.g., ~500k rows)
sampled <- ds %>%
  filter(payment_type == 1) %>%                     # ONLY credit-card trips have reliable tips
  select(tpep_pickup_datetime, tpep_dropoff_datetime,
         trip_distance, passenger_count, PULocationID,
         DOLocationID, total_amount, tip_amount,
         VendorID, RatecodeID, congestion_surcharge) %>%  # add your predictors
  # Optional: stratified sampling by hour & borough (approximate via map_batches)
  # For true stratified, you could add mutate(strata = paste(hour(tpep_pickup_datetime), floor(PULocationID/100))) 
  # but random is statistically valid and fast; extend with group_by + slice_sample after a smaller collect() if needed
  map_batches(~ as.data.frame(.) %>% 
                sample_frac(0.01)) %>%               # tweak 0.01 until you get ~200k-750k rows
  collect() %>%
  mutate(
    tip_pct = tip_amount / (total_amount - tip_amount + 1e-6),  # avoid div-by-zero
    pickup_hour = hour(tpep_pickup_datetime),
    pickup_dayofweek = wday(tpep_pickup_datetime, label = TRUE),
    trip_duration_min = as.numeric(difftime(tpep_dropoff_datetime, tpep_pickup_datetime, units = "mins")),
    # Add any other engineered features here (e.g., borough via lookup table on PULocationID)
    across(c(trip_distance, passenger_count, total_amount), ~ na_if(., 0))  # clean zeros
  ) %>%
  filter(trip_duration_min > 0, trip_distance > 0)   # basic cleaning

# Quick size check
nrow(sampled)  # should be your target sample size

# 3. Power analysis for multiple linear regression
#    First, run a pilot model on your sample (or a smaller subsample) to estimate effect size
pilot_model <- lm(tip_pct ~ trip_distance + passenger_count + pickup_hour + 
                  pickup_dayofweek + total_amount + congestion_surcharge + 
                  # ... add your other predictors
                  VendorID, data = sampled)

r_squared <- summary(pilot_model)$r.squared
f2 <- r_squared / (1 - r_squared)          # Cohen's f² from observed R²
k <- length(coef(pilot_model)) - 1         # number of predictors (u)

# Solve for required sample size given desired power = 0.80, α = 0.05
power_result <- pwr.f2.test(u = k,               # predictors
                            f2 = f2,             # or use Cohen guidelines: 0.02=small, 0.15=medium, 0.35=large
                            sig.level = 0.05,
                            power = 0.80)
required_n <- power_result$v + k + 1           # total N = v + u + 1

cat("Pilot R² =", round(r_squared, 4), 
    "\nCohen's f² =", round(f2, 3),
    "\nRequired N for 80% power with", k, "predictors =", round(required_n), "\n")

# If your actual sample is larger than required_n, you're in great shape!