---
title: "Data Cleaning Worksheet"
description: "Exploratory data analysis and cleaning steps for the NYC Yellow Taxi dataset."
author: "Aaron Powell"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    df-print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(here)

# Source utility functions using project root (fixes path issues when rendering from docs/)
source(here::here("code", "load_data.R"))
source(here::here("code", "validate_data.R"))
source(here::here("code", "clean_data.R"))
```

# Introduction

This worksheet attempts to perform exploratory data analysis and cleaning steps on the NYC Yellow Taxi dataset. Goals include identifying missing values, outliers, and inconsistencies in the data, and apply appropriate cleaning techniques to prepare the dataset for analysis. The results will be used to create a script that can be applied to the full dataset.

```{r load-data}
sample = sample_taxi_data() # starting with 750,000 observations
validate_sample(sample)
```

Our sample dataset does not significantly differ from the full dataset and appear to represent the population well judging from summary statistics. The only possible difference is mean_trip_distance, which could be influenced by lower end outliers, such as trips with 0 distance, in the population. 

# Missing Values

There are 179,044 observations with NA or missing values in the RatecodeID, passenger_count, store_and_fwd_flag, 
congestion_surcharge, airport_fee, which is approximately 2.4% of the sample. 

```{r}
missing_values = sample %>%
  filter(is.na(RatecodeID) | is.na(passenger_count) | is.na(store_and_fwd_flag) | is.na(congestion_surcharge) | is.na(Airport_fee))
dim(missing_values)
```

They appear to all be the same observations. 

```{r}
set.seed(6306)

non_missing <- sample |> 
  filter(!is.na(passenger_count))

missing_sample <- missing_values |> 
  slice_sample(n = 100)

non_missing_sample <- non_missing |> 
  slice_sample(n = 100)

var.test(
  missing_sample$total_amount, 
  non_missing_sample$total_amount
)

t.test(
  missing_sample$total_amount, 
  non_missing_sample$total_amount
)
```

Using a 99% confidence interval, there is insufficient evidence to suggest that the mean total amount of trips with missing values is different from the mean total amount of trips without missing values (welch's t-test, p-value = .015). 

```{r}
var.test(
  missing_sample$tip_amount, 
  non_missing_sample$tip_amount
)

t.test(
  missing_sample$tip_amount, 
  non_missing_sample$tip_amount
)
```

There is significant evidence to suggest that the mean tip amount of trips with missing values is different from the mean tip amount of trips without missing values (welch's t-test, p-value < .0001). The mean tip amount for trips with missing values is approximately $0.36, while the mean tip amount for trips without missing values is approximately $3.13. 

payment_type = 0 on all observations with missing values, which is flex_fare. 

check if non_missing_values have payment_type = 0
```{r}
non_missing %>% filter(payment_type == 0) %>% dim()
```

Flex fare payment types did not record values for the RatecodeID, passenger_count, store_and_fwd_flag, congestion_surcharge, and airport_fee variables. These variables are not due to data entry errors. Flex flare is an upfront payment option used by third party applications to display taxi fares to customers. These customers may be looking for budget pricing, or the third party application may not promote tipping as much as in person transactions in some ways resulting in the lower mean tip amount.

We'll create a RatecodeID for flexfare transactions as an indicator variable, and impute the other missing values with 0 since they are not applicable to flex fare transactions.

# Quality Issues

Running through a few quality checks.

```{r}
sample |> 
  summarize(
    n_leq_zero = sum(trip_distance <= 0, na.rm = TRUE),
    total = n(),
    proportion = n_leq_zero / total
  )
```

Trip distance has 21,648 observations with values less than or equal to 0. Unrealistic

```{r}
sample |> 
  summarize(
    n_leq_zero = sum(total_amount <= 0, na.rm = TRUE),
    total = n(),
    proportion = n_leq_zero / total
  )
```

Total amount has 14,989 observations with values less than or equal to 0. Unrealistic for a paid trip.

```{r}
sample |> 
  summarize(
    n_leq_zero = sum(fare_amount < 0, na.rm = TRUE),
    total = n(),
    proportion = n_leq_zero / total
  )
```

Fare amount has 48,837 observations with values less than 0. Does not match fare details on https://www.nyc.gov/site/tlc/passengers/taxi-fare.page

quick check to see if these are all flex fare transactions with missing values in RatecodeID

```{r}
sample |> 
  filter(fare_amount < 0) |> 
  summarize(
    n_missing_ratecode = sum(is.na(RatecodeID)),
    total = n()
  )
```

Roughly 3/4 of the observations with negative fare amounts are flex fare transactions with missing values in RatecodeID. We'll consider leaving these values in and use fare_amount with caution for predictions.

```{r}
sample |> 
  summarize(
    n_leq_zero = sum(passenger_count <= 0 | passenger_count > 6, na.rm = TRUE),
    total = n(),
    proportion = n_leq_zero / total
  )
```

Passenger count has 3,947 observations with values less than or equal to 0. 2 observations are greater than 6. 1,553 with 6 passengers. Target range is 1-6 passengers. 

```{r}
sample |> 
  summarize(
    n_leq_zero = sum(tip_amount < 0, na.rm = TRUE),
    total = n(),
    proportion = n_leq_zero / total
  )
```

26 observations with tip amount less than 0. Unrealistic.

Congestion surcharge, airport fee, cbd_congestion_fee are fixed values so we will use an indicator for modeling. 

# Testing data import
```{r}
cleaned_sample = clean_taxi_data(sample)
dim(cleaned_sample)
summary(cleaned_sample)
```

# Payment type review 

We'll sample from this dataset to create a manageable subset for our analysis using stratified random sampling to ensure we have all types of trips represented.

One confounding variable in tip anaylsis is cash or credit. Cash is largely determined by how much a passanger has in their pocket and we cannot assume it is reliably recorded due taxes prior to July 4, 2025 and manual entry methods. we will assume cash tipping is too confounding to model. 

We will pull two samples to justify our assumption and continue our analysis. One including cash tips and one with only credit card tips.

QOI: Is mean tip percentage significantly different between cash, credit card, or flex_rate trips?

```{r}
set.seed(6306)
sampled_for_test <- cleaned_sample %>%
  filter(payment_type %in% c(1, 2)) %>%
  group_by(payment_type) %>%
  slice_sample(n = 100) %>%
  ungroup()

t.test(tip_amount ~ payment_type, data = sampled_for_test)

t.test(tip_pct ~ payment_type, data = sampled_for_test)

mean(cleaned_sample$tip_amount[cleaned_sample$payment_type == 0], na.rm = TRUE)
mean(cleaned_sample$tip_amount[cleaned_sample$payment_type == 1], na.rm = TRUE)
mean(cleaned_sample$tip_amount[cleaned_sample$payment_type == 2], na.rm = TRUE)
```

While there is evidence the mean tip amount for cash is different than credit (p-value < .0001), cash tips are confounded by manual entry and tax polity. 

While there is evidence to suggest the mean tip percentage for flex_fare is difference than credit (welche's t-test, p-value < .0001) there is not enough information to rule out confounding variables such as the type of customer using flex fare and the third party application they are using. 

Cash has a near zero mean tip amount, while credit card tips have a mean tip amount of approximately $4.30 and flex fare tips have a mean tip amount of approximately $0.39. 

We will focus on credit card tips for the rest of our analysis due to the confounding variables in cash and flex fare tips.

```{r}
rush_true <- cleaned_sample |> filter(rush_hour)
rush_false <- cleaned_sample |> filter(!rush_hour)

var.test(rush_true$tip_amount, rush_false$tip_amount)

t.test(rush_true$tip_amount, rush_false$tip_amount)
```
There's a statistically significant difference in mean tip amounts between rush hour and non-rush hour trips (Welch's t-test p < 0.001).

```{r}
cleaned_sample |> 
  summarize(
    count_below = sum(trip_duration_min < mean(trip_duration_min)),
    total = n(),
    pct = count_below / total * 100
  )

below_avg <- cleaned_sample |> filter(trip_duration_min < mean(trip_duration_min))
above_avg <- cleaned_sample |> filter(trip_duration_min >= mean(trip_duration_min))

t.test(below_avg$tip_amount, above_avg$tip_amount)
t.test(below_avg$tip_amount, cleaned_sample$tip_amount)

standard <- cleaned_sample |> filter(RatecodeID == "Standard")
others <- cleaned_sample |> filter(RatecodeID != "Standard")

var.test(standard$tip_amount, others$tip_amount)

t.test(standard$tip_amount, others$tip_amount)

```