---
title: "EDA Notes"
description: "Notes and reflections"
author: "Aaron Powell"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    df-print: kable
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(car)
```

```{r}
source("code/load_data.R")
source("code/validate_data.R")
source("code/clean_data.R")
```

# Load the data

```{r load-data}
ds = sample_taxi_data(5000)
validate_sample(ds)
ds = clean_taxi_data(ds)
```

# Predicting Tipping Behavior

Can tipping behavior in NYC yellow taxi credit card transactions be predicted based on trip characteristics using linear regression? 

partitian data

```{r partition-data}
# Set seed for reproducibility
set.seed(6306)

n_rows <- nrow(ds)
train_idx <- sample(n_rows, size = round(0.75 * n_rows))

train <- ds[train_idx, ]
test  <- ds[-train_idx, ]
```

Review the predictor variables.

```{r}
# Review the predictor variables

# Identify numeric and categorical predictors
response <- "tip_amount"

num_vars <- train |> 
  select(where(is.numeric)) |> 
  names() |> 
  setdiff(response)

cat_vars <- train |> 
  select(where(~is.factor(.) | is.character(.))) |> 
  names()

# Sample for scatterplots (large dataset)
set.seed(6306)
train_sample <- train |> slice_sample(n = 5000)

# Function for numeric predictors - scatterplots vs tip_amount
plot_numeric <- function(var_name) {
  train_sample |> 
    ggplot(aes(x = .data[[var_name]], y = tip_amount)) +
    geom_point(alpha = 0.12, size = 0.8) +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    labs(
      title = paste(var_name, "vs tip_amount"),
      x = var_name,
      y = "Tip Amount ($)"
    ) +
    theme_bw(base_size = 11)
}

# Function for categorical predictors - boxplots
plot_categorical <- function(var_name) {
  train |> 
    ggplot(aes(
      x = reorder(.data[[var_name]], tip_amount, FUN = median),
      y = tip_amount
    )) +
    geom_boxplot(outlier.alpha = 0.5) +
    labs(
      title = paste("tip_amount by", var_name),
      subtitle = paste("Levels:", length(unique(train[[var_name]]))),
      x = NULL,
      y = "Tip Amount ($)"
    ) +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Generate and display numeric scatterplots
num_plots <- map(num_vars, plot_numeric)
wrap_plots(num_plots, ncol = 3)

# Generate and display categorical boxplots (in batches of 6)
cat_plots <- map(cat_vars, plot_categorical)
wrap_plots(cat_plots[1:6], ncol = 2)
wrap_plots(cat_plots[7:length(cat_plots)], ncol = 2)
```

trip speed and trip fare look linearily related. Trip speed is derived from trip distance and duration. 

```{r}
full = lm(tip_amount ~ trip_speed_mph, data = test)
reduced = lm(tip_amount ~ trip_distance + trip_duration_min, data = test)

anova(full, reduced)
vif(reduced)
```

Collinearity between trip_distance and trip_duration_min is moderate and does not justify the significance of the reduction in RSS (partial f-test, p-value < .0001)

```{r}
full = lm(tip_amount ~ total_amount, data = test)
reduced = lm(tip_amount ~ trip_distance + trip_duration_min, data = test)

anova(full, reduced)
vif(reduced)
```

total_amount explains more of the variance in tip_amount. Base SLR model could be total_amount and run additional models without total_amount for insights. 

```{r}
slr = lm(tip_amount ~ total_amount, data = ds)
summary(slr)

par(mfrow = c(2, 2))
plot(slr)
par(mfrow = c(1, 1))

# Histogram with normal distribution overlaid
hist(slr$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(slr$residuals), sd = sd(slr$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

using tip_pct to address non-constant variance and log(0) is imaginary. 

Attempting stepwise with AIC selection (mallow's cp)

```{r}
# Stepwise Model Selection: Forward vs Backward
library(MASS)

# Define reasonable candidate predictors
predictors <- c(
  "trip_distance", "trip_duration_min", "pickup_hour", "pickup_dow", 
  "pickup_month", "weekend", "passenger_count", "fare_per_mile", "surcharge_total", 
  "rush_hour", 'VendorID', "RatecodeID", 'trip_speed_mph', 'cbd_congestion', 'VendorID',
  "overnight", "PULocationID", "DOLocationID"
)

full_formula <- as.formula(
  paste("tip_amount ~", paste(predictors, collapse = " + "))
)

# Fit full and null models on training data
full_model <- lm(full_formula, data = train)
null_model <- lm(tip_amount ~ 1, data = train)

# Backward selection (start with full model, remove variables)
backward_model <- stepAIC(
  full_model, 
  direction = "backward", 
  trace = 1,          # set to 1 to see steps
  k = log(nrow(train)) # BIC penalty (stricter than AIC)
)

# Forward selection (start with null model, add variables)
forward_model <- stepAIC(
  null_model,
  scope = list(upper = full_formula, lower = null_model),
  direction = "forward",
  trace = 1,
  k = log(nrow(train))
)

# Compare selected models
cat("Backward selection formula:\n")
print(formula(backward_model))

cat("\nForward selection formula:\n")
print(formula(forward_model))

cat("\nBackward AIC:", AIC(backward_model))
cat("\nForward AIC: ", AIC(forward_model))
vif(backward_model)
vif(forward_model)
```

both forward and backwards now results the same with minor concerns about collinearity. 

```{r}
par(mfrow = c(2, 2))
plot(forward_model)
par(mfrow = c(1,1))

# Histogram with normal distribution overlaid
hist(forward_model$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(forward_model$residuals), sd = sd(forward_model$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

```{r}
mlr1 = lm(tip_amount ~ trip_distance + trip_duration_min + RatecodeID, data = train)
summary(mlr1)

par(mfrow = c(2, 2))
plot(mlr1)
par(mfrow = c(1,1))

# Histogram with normal distribution overlaid
hist(mlr1$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(mlr1$residuals), sd = sd(mlr1$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)

vif(mlr1)
```

adding any interactions bring multicollinearity issues. Residual plots evidence of long tailed distribution indicatng normality concerns. 

```{r}
ggplot(train, aes(x = trip_distance, y = tip_amount)) +
  geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(
    title = "Trip Distance vs Tip Amount",
    x = "Trip Distance (miles)",
    y = "Tip Amount ($)"
  ) +
  theme_bw(base_size = 18)
```

Near good looking trend line with some notable zero tip values. 

```{r}
ggplot(train, aes(x = trip_duration_min, y = tip_amount)) +
  geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(
    title = "Trip Distance vs Tip Amount",
    x = "Trip Distance (miles)",
    y = "Tip Amount ($)"
  ) +
  theme_bw(base_size = 18)
```

Trip distance may be skewed due to tip amount being zero at large distances. let's test what the relationship looks like without zero values.

```{r}
ggplot(train, aes(x = trip_duration_min, y = tip_amount)) +
  geom_point(alpha = 0.1, size = 0.8) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(
    title = "Trip Distance vs Tip Amount",
    x = "Trip Distance (miles)",
    y = "Tip Amount ($)"
  ) +
  theme_bw(base_size = 18)
```

```{r}
summary(train$tip_amount)
hist(train$tip_amount, breaks=100) 
```

Very large amount of zero tip values. 

Compare rmse mlr1 to a full model of predictors 

```{r}
library(car)
full_mod <- lm(tip_amount ~ trip_distance + trip_duration_min + RatecodeID + 
               pickup_hour + pickup_dow + weekend + rush_hour + 
               airport + cbd_congestion + fare_per_mile + log_total_amount + 
               passenger_count + VendorID, 
               data = train)

summary(full_mod)
```

good for some insights. looking to add one more variable to get above 60 adjusted r^2

Further investigate why rush_hour did not make it into the model. rush_hour is entangled with trip_distance and trip_duration_min, but by how much? compare rmse mlr1 to a model with rush_hour.

```{r}
mlr2 = lm(tip_amount ~ trip_distance + trip_duration_min + PULocationID + DOLocationID, data = train)
summary(mlr2)
anova(mlr2, mlr1)

par(mfrow = c(2, 2))
plot(mlr2)
par(mfrow = c(1,1))

# Histogram with normal distribution overlaid
hist(mlr2$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(mlr2$residuals), sd = sd(mlr2$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

mlr2 better predictive power but more complex and residual plots are not great. Will include in final evaluation.

```{r}
mlr3 = lm(tip_amount ~ trip_distance + trip_duration_min + pickup_hour + pickup_dow + pickup_month, data = train)
summary(mlr3)
anova(mlr3, mlr2)

par(mfrow = c(2, 2))
plot(mlr3)
par(mfrow = c(1,1))

# Histogram with normal distribution overlaid
hist(mlr3$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(mlr3$residuals), sd = sd(mlr3$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

While there is evidence pickup time is significant, the f-statistic is less than mlr2 using location predictors.

# Is tipping dependent on time of day, day of week, or month?

