---
title: "Predictions"
description: "Modeling process for the NYC Yellow Taxi dataset."
author: "Aaron Powell"
date: today
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    code-tools: true
    df-print: kable
---

```{r setup, include=FALSE}
library(tidyverse)
library(car)
library(broom)
source("code/load_data.R")
source("code/validate_data.R")
source("code/clean_data.R")
set.seed(6306)
```

# Data

```{r}
ds = sample_taxi_data(n=5000)
validate_sample(ds)
ds = clean_taxi_data(ds)
train_idx = sample(nrow(ds), size = round(0.75 * nrow(ds)))
train = ds[train_idx, ]
test = ds[-train_idx, ]
```

# Feature Selection

Can tipping behavior in NYC yellow taxi credit card transactions be predicted based on trip characteristics using linear regression? 

During exploratory analysis, scatterplots and boxplots were reviewed for linear relationships against tip_amount. Further investigation revealed total_amount to be the best statistical predictor. Total_amount is a function of all other variables, but mostly trip_distance and trip_duration_min. While we could use total_amount as the single best predictor it does not provide us with practical prediction for customers, executives or investors. Customers know the average tip amount is 17% and will likely tip between 10-20%. Executives want to ensure their employees are supporting in a thriving business of taxi service. Investors need insights. These models will focus on a predictive model for executives which could be applied for business analytics such as compensation reviews or contractor negotiations.

```{r model creation}
predictors = c(
    "trip_distance", "trip_duration_min", "pickup_hour", "pickup_dow", 
    "pickup_month", "weekend", "passenger_count", "fare_per_mile", "surcharge_total", 
    "rush_hour", 'VendorID', "RatecodeID", 'trip_speed_mph', 'cbd_congestion', 'VendorID',
    "overnight" # Leaving out PUlocationID and DOLocationID. Unable to find full list of IDs for factoring.
    )
# Exemplify overfit with full model
full_formula <- as.formula(paste("tip_amount ~", paste(predictors, collapse = " + ")))
full_model <- lm(full_formula, data = train)

# Simpliest model
slr = lm(tip_amount ~ total_amount, data = train)

library(MASS)     # stepAIC this library select() method conflicts with dplyr

# Best parsimonious fit
a_model <- stepAIC(
    full_model, 
    direction = "both", 
    trace = 0, 
    k = log(nrow(train)))  # BIC for parsimony

library(dplyr) # reload correct library

# Exemplify adding next best indicator to stepwise solution
b_model = lm(
    tip_amount ~ trip_distance + trip_duration_min + RatecodeID + rush_hour,
    data= train)
```

# Diagnostics

Residual plots are review for ensuring linear regression is robust enough to perform well as prediction model. In other words, does the shoe fit?

```{r model residual plots}
par(mfrow = c(2, 2))
plot(full_model)
plot(slr)
plot(a_model)
par(mfrow = c(1,1))

# Histogram with normal distribution overlaid
hist(full_model$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals (Full Model)",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(full_model$residuals), sd = sd(full_model$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)

# Histogram with normal distribution overlaid
hist(slr$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals (Simple Model)",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(slr$residuals), sd = sd(slr$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)

# Histogram with normal distribution overlaid
hist(a_model$residuals, 
     probability = TRUE,          # Use density scale
     main = "Histogram of Residuals (A Model)",
     xlab = "Residuals",
     col = "lightblue",
     border = "black")
# Add normal curve
curve(dnorm(x, mean = mean(a_model$residuals), sd = sd(a_model$residuals)), 
      add = TRUE, 
      col = "red", 
      lwd = 2)
```

Results are not perfect, there's some skewness in normality and zero tip annotated in the dataset are only slightly in disagreeance with assuming a constant variance among predictor variables. Outliers have been appropriately addressed while leaving room for influence in the model.

# Comparison

Model A provides the best explaination of variance while minimizing complexity. Adding all predictors would only explain 8% more of the variance, while adding the next best predictor would not explain even .1% more. The trade-off is 6 degrees of freedom less than the simpliest model.

```{r model-comparison}
# Create comparison table from model summaries
comparison_table <- map_dfr(list(
    "Full Model" = full_model,
    "SLR (total_amount)" = slr,
    "Stepwise BIC (a_model)" = a_model,
    "Manual (b_model)" = b_model
  ), ~ {
    g <- broom::glance(.x)
    tibble(
      Model = names(which(map_lgl(list(full_model, slr, a_model, b_model), identical, .x))),
      Observations = g$nobs,
      `R²` = round(g$r.squared, 4),
      `Adj. R²` = round(g$adj.r.squared, 4),
      `Residual SE` = round(g$sigma, 3),
      `F-statistic` = round(g$statistic, 2),
      `p-value` = format.pval(g$p.value, digits = 3, eps = 0.001),
      `Residual Df` = g$df.residual
    )
  }, .id = NULL)

comparison_table |>
  knitr::kable(
    caption = "Model Summary Comparison",
    align = "lrrrrrr"
  )
```

There is some expected collinearity between total_distance and total_duration_min but minimal for the dataset. 

```{r vif-comparison}
# VIF (Variance Inflation Factor) Comparison
# Note: Full model is too complex for standard VIF due to location IDs (277 predictors)

library(car)

# Named list of models
models = list(
    "Full Model" = full_model,
    "SLR (total_amount)" = slr,
    "Stepwise BIC (a_model)" = a_model,
    "Manual (b_model)" = b_model
    )

# Calculate VIF where feasible
get_vif_summary <- function(model, name) {
  if (name == "Full Model") {
    return(tibble(
      Model = name,
      `Max VIF` = NA_real_,
      `Mean VIF` = NA_real_,
      `VIF > 5` = NA_integer_,
      `VIF > 10` = NA_integer_,
      `Total Predictors` = length(coef(model)) - 1,
      Notes = "Too complex (includes PULocationID + DOLocationID)"
    ))
  }
  
  vifs <- tryCatch({
    vif_values <- vif(model)
    if (is.matrix(vif_values)) {
      # For models with categorical variables (GVIF)
      vif_values[, "GVIF^(1/(2*Df))"]
    } else {
      vif_values
    }
  }, error = function(e) rep(1, length(coef(model)) - 1))
  
  tibble(
    Model = name,
    `Max VIF` = round(max(vifs, na.rm = TRUE), 2),
    `Mean VIF` = round(mean(vifs, na.rm = TRUE), 2),
    `VIF > 5` = sum(vifs > 5, na.rm = TRUE),
    `VIF > 10` = sum(vifs > 10, na.rm = TRUE),
    `Total Predictors` = length(vifs),
    Notes = ""
  )
}

vif_table <- map_dfr(names(models), ~ get_vif_summary(models[[.x]], .x))

vif_table |>
  knitr::kable(
    caption = "VIF Comparison Across Models",
    align = "lrrrrrc",
    digits = 2
  )
```

# Performance

Train Adj. R²: In-sample fit (adjusted for predictors)
Test RMSE / MAE: How well the model predicts tip_amount on unseen data
Test R²: Out-of-sample explanatory power
AIC/BIC: Model selection criteria

```{r model-performance}
# Generate predictions on held-out test set
test_preds <- test |>
  mutate(
    pred_full = predict(full_model, newdata = test),
    pred_slr  = predict(slr, newdata = test),
    pred_a    = predict(a_model, newdata = test),
    pred_b    = predict(b_model, newdata = test)
  )

# Helper functions for test set evaluation
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

mae <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

rsq <- function(actual, predicted) {
  1 - sum((actual - predicted)^2, na.rm = TRUE) / 
      sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
}

# Comprehensive model evaluation table
model_evaluation <- tibble(
  Model = c("Full Model", 
           "SLR (total_amount)", 
           "Stepwise BIC (a_model)", 
           "Manual (b_model)"),
  
  `Train Adj. R²` = c(
    round(summary(full_model)$adj.r.squared, 4),
    round(summary(slr)$adj.r.squared, 4),
    round(summary(a_model)$adj.r.squared, 4),
    round(summary(b_model)$adj.r.squared, 4)
  ),
  
  `Test RMSE` = c(
    round(rmse(test$tip_amount, test_preds$pred_full), 3),
    round(rmse(test$tip_amount, test_preds$pred_slr), 3),
    round(rmse(test$tip_amount, test_preds$pred_a), 3),
    round(rmse(test$tip_amount, test_preds$pred_b), 3)
  ),
  
  `Test MAE` = c(
    round(mae(test$tip_amount, test_preds$pred_full), 3),
    round(mae(test$tip_amount, test_preds$pred_slr), 3),
    round(mae(test$tip_amount, test_preds$pred_a), 3),
    round(mae(test$tip_amount, test_preds$pred_b), 3)
  ),
  
  `Test R²` = c(
    round(rsq(test$tip_amount, test_preds$pred_full), 4),
    round(rsq(test$tip_amount, test_preds$pred_slr), 4),
    round(rsq(test$tip_amount, test_preds$pred_a), 4),
    round(rsq(test$tip_amount, test_preds$pred_b), 4)
  ),
  
  AIC = c(
    round(AIC(full_model), 1),
    round(AIC(slr), 1),
    round(AIC(a_model), 1),
    round(AIC(b_model), 1)
  ),
  
  BIC = c(
    round(BIC(full_model), 1),
    round(BIC(slr), 1),
    round(BIC(a_model), 1),
    round(BIC(b_model), 1)
  )
)

model_evaluation |>
  knitr::kable(
    caption = "Model Performance Evaluation (Training + Held-Out Test Set)",
    align = "lrrrrrr"
  )
```

```{r test-prediction-summary}
# Summary of test set prediction accuracy
summary_stats <- test_preds |>
  summarize(
    Mean_Actual = mean(tip_amount),
    Mean_Predicted_Full = mean(pred_full),
    Mean_Predicted_SLR = mean(pred_slr),
    Mean_Predicted_B = mean(pred_b),
    .groups = "drop"
  )

summary_stats |> knitr::kable(
  caption = "Mean Predicted vs Actual Tip Amount on Test Set",
  digits = 3
)
```

